Select-based random access to variable-byte encodings

### 2020-03-02 gradu/msc_english_contents.tex
Changes

$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better.

@@

For example, if block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
\texttt{00000000 00101000}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1000}. The block with least significant bits is given a continuation bit 1.
The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11000}, 
which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}.

@@

\section{VB decoding}
VB decoding is technically just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original number. An example implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.

As in the encoding example, block length is 4, the encoded message 
is \texttt{00010 11000} and n is a 16 bit unsigned integer. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
$n = $\texttt{101000} (bit representation of 42). Because the continuation bit of this chunk is 1, decoding for this number is complete.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[H]
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State $numbers$.append($n$)
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\end{minipage}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

@@

\begin{table}
\centering
\caption{Memory requirements of SDSL rank and select data structures\label{table:supportsize}}
\begin{tabular}{l||c} 
Structure & Extra size taken\\ 
\hline \hline 
Rank   & \text{25\% of bit array} \\
Rankv5 & \text{6.25\% of bit array}\\
Select & \text{~8.3\% of bit array (on average)}\\
\hline
\end{tabular}
\end{table}

@@

Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
library \citep{gbmp2014sea}. Each implementation is decoding a uint64\_t integer.

@@

Five differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).

\begin{itemize}
  \item all - all integers randomly 1-4 bytes long
  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
  \item twolarge2 - Same as twolarge, but third of the 1 byte integers 4 bits or less
  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
  \item onebyte - all integers 4 bits or less
\end{itemize}

All implementations read the data set from a file and compress it to memory. Then one million index integers are randomized and stored to an array. The times shown in Table~\ref{table:results1} 
are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.

@@

 - Run with 32 bit values, see what happens (some operations may be 32b?)
 -

>>>>

$Rank$ method is superior when dataset contains mostly small integers. With larger integers, multiple $rank$ calls are needed and $select$ method performs better. Additionally, the
proposed $select$ method offers quick access to next or previous element.

@@

For example, block length is set to 4 and 42 is a 16 bit unsigned integer that is to be encoded. The standard 8-bit representation of 42 is 
\texttt{00000000 00101000}. This is then split to blocks of 4 bits, resulting in \texttt{0000 0000 0010 1000}. The block with least significant bits is given a continuation bit 1.
The block with the next least significant bits is given a continuation bit 0. The rest of the blocks are omitted, because they all are empty. The result of this is \texttt{00010 11000}, 
which is the number 42 compressed with VB encoding and block length of 4. A pseudo code with block length 4 is shown in Figure~\ref{vbyte_enc}. Prepend adds an element to the beginning of the list 
and extend adds all the elements of the second list to the end of the first list.

@@

\section{VB decoding}
VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original integer. A pseudo code implementation of VB decoding with block length of 7 is shown in Figure~\ref{vbyte_dec}.
Append adds an element to the end of the list. If block length is changed, additional bit operation steps when reading the data may be needed.

As how the encoding example ended, the encoded message was \texttt{00010 11000} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
extracted and added to $n$, making $n = $\texttt{10} (bit representation of 2).
The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
$n = $\texttt{101000} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[H]
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State \Call{append}{numbers,n}
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\end{minipage}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

@@

\begin{table}
\centering
\caption{Memory requirements of SDSL rank and select data structures\label{table:supportsize}}
\begin{tabular}{l||c} 
Structure & Extra size taken\\ 
\hline \hline 
Rank   & \text{25\% of bit array} \\
Rankv5 & \text{6.25\% of bit array}\\
Select & \text{~8-12\% of bit array}\\
\hline
\end{tabular}
\end{table}

@@

Five different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t integers.

@@

Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).

\begin{itemize}
  \item all - all integers randomly 1-4 bytes long
  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
  \item onlysmall - all integers 4 bits or less
\end{itemize}

All implementations read the data set from a file and compress it to memory. Then one million index integers are randomized and stored to an array. The times shown in Table~\ref{table:results1} 
are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated multiple times and an average of the runtimes is taken.

@@

 - get compression sizes of 4 and 8 bit versions

 - figure out why data size matters
